### 2022-06-14 manuscript/hiv-gwas.tex
Update on Overleaf.

\documentclass[11pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{lineno} % for line numbers
\usepackage{graphicx}
\usepackage{bm} % for bold math characters
\usepackage{booktabs} % for prettier hlines in table
\usepackage{subcaption} % for subfigures
\usepackage[affil-it]{authblk} % for footnote-style author affiliation
\usepackage{setspace} % for 1.5-spaced lines not including footnotes, floats
\usepackage{tikz} % for graphical model drawing
\usepackage{float} % to enforce figure positioning in subsection they're declared in
\usepackage{amsmath}
\usepackage[singlelinecheck=false]{caption} % to left-justify single-line table captions
\usepackage{array}
\newcolumntype{P}[1]{>{\centering\arraybackslash}p{#1}} % for centered, width-specified columns 
\usepackage{makecell} % for multi-row table entries
\usepackage{tabularx} % for tables that don't spill past margin

@@

\begin{abstract}
	Infectious diseases are particularly challenging for genome-wide association studies (GWAS) because genetic effects from two organisms (pathogen and host) can influence a trait. Traditional GWAS assume individual samples are independent observations. However, pathogen effects on an infectious disease trait can be heritable from donor to recipient in transmission chains. Thus, residuals in GWAS association tests for host genetic effects may not be independent due to shared pathogen ancestry. We propose a new method to estimate and remove heritable pathogen effects on a trait based on the pathogen phylogeny prior to host GWAS, thus restoring independence of samples. In simulations, we show this pre-processing step can increase GWAS power to detect truly associated host variants when pathogen effects are highly heritable, with strong phylogenetic correlations. When we apply our method to HIV-1 subtype B data from the Swiss HIV Cohort Study, we recover slightly weaker but qualitatively similar signals of association between set-point viral load (spVL) and human genetic variants in the \emph{CCR5} and major histocompatibility complex (MHC) gene regions compared to standard GWAS. Our simulation study confirms that based on the estimated heritability and selection parameters for HIV-1 subtype B spVL, standard GWAS are robust to pathogen effects. Then, we show that phylogenetic correlations for quantitative disease resistance (QDR) in the \emph{A. thaliana}-\emph{X. arboricola} pathosystem are not strong enough to justify a phylogenetic correction for GWAS. Our framework may improve GWAS for other diseases if pathogen effects are even more phylogenetically correlated amongst individuals in a cohort.
\end{abstract}

@@

Variation in infectious disease traits like viral load or infection severity can come from several sources. These include host genetic factors, pathogen genetic factors, interaction effects between the host and the pathogen, or non-genetic factors like healthcare quality or temperature. GWAS typically stratify samples or include covariates to correct for host genetic factors or non-genetic factors that may be correlated with a trait value. This leaves pathogen genetic factors as a remaining source of correlation, since close transmission partners may be infected with very similar pathogen strains. Broad-sense heritability $H^2$ quantifies the fraction of total trait variance that is ``inherited'' from infection partner to infection partner. To characterize $H^2$ and the heritable and non-heritable factors that determine infectious disease traits, we use a phylogenetic mixed model (PMM) \citep{Housworth2004}. PMMs assume continuous traits are the sum of independent heritable and non-heritable parts. In the infectious disease GWAS case, we assume the heritable part comprises pathogen genetic factors and all other factors are non-heritable. The heritable pathogen part is modeled by a random process occurring in continuous time along the branches of the pathogen phylogeny, as in Figure \ref{fig:spVLModel}A. The non-heritable part is modeled as Gaussian noise added to sampled individuals at the tips of the phylogeny.

@@

First, we evaluated how well our method estimated the additive host genetic effects from the simulated data. Additive host genetic effects represent an ideal (albeit unattainable) baseline for infectious disease GWAS. Figure \ref{fig:simulationResults}A shows that our method incorporating phylogenetic information can more accurately estimate these value compared to the trait value. To ensure a fair comparison, we scaled trait values to have the same mean, zero, as host genetic effects so as not to bias the root mean squared error (RMSE) by a constant factor. As shown in the supplemental material, we can calculate the expected RMSE using the scaled trait value across scenarios in our simulation scheme because the variance in the trait due to viral genetic effects and environmental effects is fixed. Thus, we expect the RMSE using the scaled trait value to be 0.74 across all simulation scenarios. By incorporating phylogenetic information, we can improve upon this error in scenarios where the trait is highly heritable, under low selection pressure, and with relatively moderate stochastic fluctuations compared to outbreak duration. Figure \ref{fig:simulationTraitExample} gives some intuition for how this correction works by contrasting simulated scenarios with high and low heritability and low selection strength. These parameter combinations mean trait values are phylogentically correlated at clustered tips in the phylogeny. This information is the basis from which our method estimates the pathogen and corresponding non-pathogen effects on the trait. This is because high heritability means the trait value is highly pathogen-dependent. Then, when the trait is under weak selection, the pathogen effects can drift far from the long-term optimum and stochastic fluctuations are low to maintain the same heritability. Thus, our method performs well when an infectious disease trait is highly heritable and trait values are strongly correlated amongst close transmission partners.

@@

\begin{figure}[H]
	\begin{center}
		\includegraphics[width=\linewidth]{figures/simulation_results}
		\caption{Results from the simulation study. We simulated host, pathogen, and environmental effects on a trait under the POUMM with different heritability (y-axis) and selection strength (x-axis) parameters. For each simulated dataset, we applied our method to estimate the non-pathogen effects and performed GWAS with these values. (A) shows that our method (left) can generate more accurate estimates of additive host genetic effects than the trait value, scaled by its mean (right). (B) shows how GWAS power can improve given the true, simulated non-pathogen effect on spVL (left) and using our estimate for this value (middle) compared to using the scaled trait value (right). Each tile's color corresponds to the average value across 20 simulated datasets of 500 samples. The points highlight specific heritability and selection strength values from the \emph{A. thaliana}-\emph{X. arboricola} QDR analysis, HIV-1 spVL analysis, and four simulated scenarios that are presented in more detail in Figure \ref{fig:tip-trait-correlation}. RMSE = Root mean square error, TPR = True positive rate, QDR = quantitative disease resistance, spVL = set-point viral load.}
		\label{fig:simulationResults}
	\end{center}
\end{figure}

\begin{figure}[H]
	\begin{center}
		\includegraphics[width=\linewidth]{figures/example_simulation_trait_values}
		\caption{Simulated data from two evolutionary scenarios where a phylogenetic correction to trait values improves GWAS power (right side) and where it does not (left side). These examples correspond to the empty points in Figure \ref{fig:simulationResults}. (A) and (B) show total trait values for 12 randomly selected tips from the simulated phylogeny with pathogen heritability 15 and 75\%, respectively. Depending on the pathogen heritability, trait values are more or less correlated at clustered tips. (C) compares our method's estimate for the non-pathogen part of trait values (y-axis) with true simulated host trait values (x-axis) with pathogen heritability 15 and 75\%. The solid line is the y=x line. Selection strength was fixed to 0.1 time $^{-1}$ for both scenarios and all other parameters were fixed as in the full simulation study.}
		\label{fig:simulationTraitExample}
	\end{center}
\end{figure}

\begin{figure}[H]
\begin{center}
\includegraphics[width = \linewidth]{figures/tip_trait_correlation.png}
	\caption{Correlations between trait values in pairs of tips in four of the simulated scenarios from Figure \ref{fig:simulationResults}. Correlations are calculated for pairs of tips binned by phylogenetic distance (into deciles) across the 20 replicate simulations for each evolutionary scenario. Trait values are only noticeably correlated for closely clustered tips under the scenario with high pathogen heritability and low selection strength (upper right facet).}
	\label{fig:tip-trait-correlation}
	\end{center}
\end{figure}

@@

We applied our framework to empirical data from two different host-pathogen systems. First, we used data collected by the Swiss HIV Cohort Study (SHCS) from 1,392 individuals in Switzerland infected with HIV-1 subtype B between 1994 and 2018. The SHCS provided viral load measurements, $pol$ gene sequences, and human genotype data for these individuals. We followed the framework outlined above to estimate the pathogen and non-pathogen effects on spVL for the cohort from these data. Figure \ref{fig:spvl-calc-comparison} shows the calculated (total) spVL values, which vary between approximately 1 and 6 log copies/mL in the cohort.  We estimated spVL heritability in this cohort to be 45\% (95\% highest posterior density, HPD, 24 - 67\%) and selection strength to be 58 time$^{-1}$ (95\% HPD 19 - 95) (Figure \ref{fig:poumm-parameters}, Table \ref{tab:POUMM-parameters-our-estimates}). To put these values into the context of our simulation study, they are shown as points on Figure \ref{fig:simulationResults}. The highest expected correlation in trait values between any two tips in the HIV-1 phylogeny under the POUMM was 0.45. However, Figure \ref{fig:spvl-on-tree} shows that this trait is not obviously phylogenetically structured in the cohort in general, despite high heritability. Finally, figure \ref{fig:trait-comparison} shows that the estimated non-pathogen effects on spVL correlate quite strongly with total spVL.

We compared our proposed GWAS framework with a more standard approach by performing two different GWAS on the same SHCS human genotypes. In the (i) ``GWAS with standard trait value'' we used the total trait value, our calculated spVL values, as the GWAS response variable. In the (ii) ``GWAS with estimated non-pathogen part of trait'' we used our estimates for the non-pathogen effects on spVL. Figure \ref{fig:gwas-results}A shows that results are qualitatively similar between the two GWAS. Q-Q plots show the distribution of p-values are very similar as well (Figure \ref{fig:qq-plots}). Figure \ref{fig:gwas-results}B shows how the strength of association changed for some variants in the MHC and \emph{CCR5} regions. Taking into account phylogenetic information slightly decreased association strength for most variants in the \emph{CCR5} region. Association strength increased for some variants in the MHC, for example, SNP rs9265880 had the greatest increase in significance in the MHC region, from a p-value of $3.5 \times 10^{-07}$ to $7.7 \times 10^{-09}$. However, the top-associated variants in the MHC and \emph{CCR5} regions were consistent regardless of the GWAS response variable used (Table \ref{tab:gwas-top-associations}). Finally Table \ref{tab:comp-gwas-mclaren-snps} shows how our GWAS results compare for the two top-associated SNPs identified by \citet{McLaren2015}. In summary, there are no clear patterns that point to new regions of association in the human genome with spVL when we take into account the pathogen phylogeny.

@@

Next, we applied our framework to data collected from the \emph{A. thaliana}-\emph{X. arboricola} pathosystem by \cite{Wang2018Two-wayGenomes}. \cite{Wang2018Two-wayGenomes} performed a fully-crossed experiment in which they infected genetically diverse \emph{A. thaliana} accessions with genetically diverse strains of the phytopathogenic bacteria \emph{X. arboricola}. They scored quantitative disease resistance (QDR) on a scale of 0 (resistant) to 4 (susceptible) for up to four infected leaves for three replicates of each \emph{A. thaliana}-\emph{X. arboricola} pairing. Figure \ref{fig:qdr-on-tree} shows the inferred \emph{X. arboricola} pathogen phylogeny. Mean QDR calculated across all \emph{A. thaliana} pairings and all replicates for each \emph{X. arboricola} strain was generally low, varying between 0.11 for strain NL\_P126 and 0.78 for strain FOR\_F21. Fitting the POUMM yielded very low selection strength $\alpha$ and intensity of stochastic fluctuations  $\sigma$ parameter estimates (posterior mean 0.03 with 95\% HPD 0.0 - 0.05 and 0.03 with 95\% HPD 0.0 - 0.06, respectively; Table \ref{tab:POUMM-parameters-our-estimates-qdr}). These values deviated significantly from the respective priors (Figure \ref{fig:poumm-parameters-qdr}). Heritability, on the other hand, was quite uncertain (posterior mean 0.33 with 95\% HPD 0.0 - 0.77; Table \ref{tab:POUMM-parameters-our-estimates-qdr}). These selection strength and heritability values are also shown in the context of the simulation study as points on Figure \ref{fig:simulationResults}.

@@

\section*{Discussion}
% State the importance of your findings
In this paper, we presented a new phylogeny-aware GWAS framework to correct for heritable pathogen effects on infectious disease traits. By using information from the pathogen phylogeny, we show it is possible to improve GWAS power to detect host genetic variants associated with a disease trait. This should help us better understand which host factors are protective against a disease versus which increase susceptibility or disease severity.

@@

The main novelty of our approach is to use these POUMM parameter estimates to correct infectious disease trait values prior to performing GWAS, thereby estimating and removing pathogen effects. In simulations, we show that when trait heritability amongst infection partners is greater than approximately 25\%, GWAS power to detect host genetic variants associated with the same trait is reduced. Our method can correct for this effect in certain evolutionary scenarios by using information from the full pathogen phylogeny. Based on our simulation results, our method is anticipated to be very useful for disease traits that are highly heritable from donor to recipient and  maintain a high correlation between sampled individuals. In simulations, we showed this is the case when heritability is high, selection strength is low, and trait values are not subject to strong stochastic fluctuations. So, cohort-level, phylogenetically structured differences in the measured trait value are necessary for our approach to outperform state of the art methods.

Given our estimates for trait heritability and selection strength on HIV-1 spVL and \emph{A. thaliana} quantitative disease resistance to \emph{X. arboricola}, our simulation results reveal that we cannot expect a significant improvement in GWAS power (Figure \ref{fig:simulationResults}). Certain pairs of samples in the HIV-1 cohort were indeed expected to have phylogenetically correlated spvL values (maximum expected correlation between any two samples was 0.45). However, the overall effect on GWAS is small. For HIV-1 spVL, our phylogenetic correction slightly decreases p-values for variants in $CCR5$ and slightly decreases some and increases other p-values for variants in the MHC (Figure \ref{fig:gwas-results}B). Simulations show we shouldn't expect a net p-value decrease, but our simulations represent an ideal scenario since we simulate under the POUMM. In real life, un-modeled evolutionary pressures like drug treatment and host-specific HLA alleles might cause the reduced p-values. However, the overall picture is consistent between the two GWAS (Figure \ref{fig:gwas-results}A). Therefore, we conclude that GWAS for host determinants of HIV-1 subtype B spVL is robust to our correction for pathogen effects. For \emph{A. thaliana} QDR, inferred phylogenetic correlations are too weak to justify a phylogenetic correction (maximum expected correlation between strains was $3.2 \times 10^{-12}$). \cite{Wang2018Two-wayGenomes} found no significant \emph{A. thaliana} variants associated with QDR using a linear mixed model jointly accounting for host genetic effects, pathogen genetic effects, and interaction effects. As with HIV-1 spVL, our results do not challenge this previous finding.

Our approach is envisioned to be complementary to previously developed methods for infectious disease GWAS. First, it provides additional information on the evolutionary dynamics of the trait in the pathogen population. Then, it is a convenient pre-processing step for GWAS because it simply produces an alternate response variable for GWAS association tests. As in the \emph{A. thaliana}-\emph{X. arboricola} application, fitting the POUMM may reveal that expected phylogenetic correlations between samples are not strong enough to justify using the corrected trait values in a GWAS. In this case, one may wish to use a linear mixed model as in \citep{Wang2018Two-wayGenomes}, where the pathogen effect is co-estimated as a random effect. In cases where a correction can be estimated and applied, it is still possible to use any of the number of previously developed GWAS models for the actual association testing (we used a linear model approach implemented in PLINK \citep{Chang2015}, though a more advanced method would be to use a linear mixed model with host ancestry as a random effect). Finally, additional model complexity can be added to the GWAS association tests. For instance, our method does not account for co-infection, although one could add co-infection status as a covariate in the GWAS association test to account for this variable.

Our method relies on the freely available R package POUMM \citep{Mitov2017a-POUMM}, which scales to trees of up to 10,000 tips \citep{Mitov2019}. All code for the simulations and HIV spVL analysis presented in this study is available on the project GitHub at https://github.com/cevo-public/POUMM-GWAS. Future applications of our method might investigate other clinically significant disease traits and outcomes that are affected by both host and pathogen genetic factors, for instance Hepatitis B Virus-related hepatocellular carcinoma \citep{An2018HostCarcinoma}, Hepatitis C treatment success \citep{Ansari2017Genome-to-genomeVirus}, and susceptibility to or severity of certain bacterial infections, e.g. \cite{Messina2016ImpactInfections, Donnenberg2015BacterialStudies}.

@@

\subsection*{\emph{A. thaliana}-\emph{X. arboricola} data}
\emph{A. thaliana} and \emph{X. arboricola} genotyping and quantitative disease resistance (QDR) measurements are described in detail in \citep{Wang2018Two-wayGenomes}. Briefly, \citep{Wang2018Two-wayGenomes} infected different \emph{A. thaliana} host accessions with different \emph{X. arboricola} pathogen strains in a fully-crossed experimental design. They infected up to 4 leaves on each of three biological replicates for each host-pathogen pair. Then, they scored QDR for each leaf on a scale of 0 (resistant) to 4 (susceptible). A genotype matrix with allele dosage of 33,610 SNPs for the 22 \emph{X. arboricola} pathogen strains used was downloaded from the supplemental material of \citep{Wang2018Two-wayGenomes}. A VCF file with allele dosage of 12,883,854 SNPs for different \emph{A. thaliana} accessions was downloaded from the 1001 Genomes project \citep{Alonso-Blanco20161135Thaliana}. QDR measurements were provided directly by the authors of \citep{Wang2018Two-wayGenomes}.

For phylogenetic inference, we used the ``dist.gene'' and ``nj'' functions from the ape package in R to construct a pairwise genetic distance matrix and then a neighbor-joining tree from the \emph{X. arboricola} pathogen genotype matrix. For the trait value to fit the POUMM, we calculated mean QDR across all leaves infected on all hosts for each \emph{X. arboricola} strain.

>>>>

\documentclass[11pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{lineno} % for line numbers
\usepackage{graphicx}
\usepackage{bm} % for bold math characters
\usepackage{booktabs} % for prettier hlines in table
\usepackage{subcaption} % for subfigures
\usepackage[affil-it]{authblk} % for footnote-style author affiliation
\usepackage{setspace} % for 1.5-spaced lines not including footnotes, floats
\usepackage{tikz} % for graphical model drawing
\usepackage{float} % to enforce figure positioning in subsection they're declared in
\usepackage{amsmath}
\usepackage[singlelinecheck=false]{caption} % to left-justify single-line table captions
\usepackage{array}
\newcolumntype{P}[1]{>{\centering\arraybackslash}p{#1}} % for centered, width-specified columns 
\usepackage{makecell} % for multi-row table entries
\usepackage{tabularx} % for tables that don't spill past margin
\usepackage[markup=underlined]{changes}

@@

\begin{abstract}
	Infectious diseases are particularly challenging for genome-wide association studies (GWAS) because genetic effects from two organisms (pathogen and host) can influence a trait. Traditional GWAS assume individual samples are independent observations. However, pathogen effects on a trait can be heritable from donor to recipient in transmission chains. Thus, residuals in GWAS association tests for host genetic effects may not be independent due to shared pathogen ancestry. We propose a new method to estimate and remove heritable pathogen effects on a trait based on the pathogen phylogeny prior to host GWAS, thus restoring independence of samples. In simulations, we show this pre-processing step can increase GWAS power to detect truly associated host variants when pathogen effects are highly heritable, with strong phylogenetic correlations. We applied our framework to data from two different host-pathogen systems. For HIV-1 set-point viral load (spVL) in the Swiss HIV Cohort Study, we recover slightly weaker but qualitatively similar signals of association between spVL and human genetic variants in the \emph{CCR5} and major histocompatibility complex (MHC) gene regions. Our simulation study confirms that based on the estimated heritability and selection parameters for HIV-1 spVL, standard GWAS are robust to pathogen effects. For \emph{A. thaliana} quantitative disease resistance (QDR) to \emph{X. arboricola}, phylogenetic correlations are not strong enough to justify our phylogenetic correction for GWAS. In summary, our framework provides additional information on the evolutionary dynamics of traits in pathogen populations and may improve GWAS if pathogen effects are highly phylogenetically correlated amongst individuals in a cohort.
\end{abstract}

@@

Variation in infectious disease traits like viral load or infection severity can come from several sources. These include host genetic factors, pathogen genetic factors, interaction effects between the host and the pathogen, or non-genetic factors like healthcare quality or temperature. GWAS typically stratify samples or include covariates to correct for host genetic factors or non-genetic factors that may be correlated with a trait value. This leaves pathogen genetic factors as a remaining source of correlation, since close transmission partners may be infected with very similar pathogen strains. Broad-sense pathogen heritability $H^2$ quantifies the fraction of total variance in a trait that is ``inherited'' from infection partner to infection partner, i.e., due to pathogen factors. To characterize $H^2$ and the heritable and non-heritable factors that determine infectious disease traits, we use a phylogenetic mixed model (PMM) \citep{Housworth2004}. PMMs assume continuous traits are the sum of independent heritable and non-heritable parts. In the infectious disease GWAS case, we assume the heritable part comprises pathogen genetic factors and all other factors are non-heritable. The heritable pathogen part is modeled by a random process occurring in continuous time along the branches of the pathogen phylogeny, as in Figure \ref{fig:spVLModel}A. The non-heritable part is modeled as Gaussian noise added to sampled individuals at the tips of the phylogeny.

@@

First, we evaluated how well our method estimated the additive host genetic effects from the simulated data. Additive host genetic effects represent an ideal (albeit unattainable) baseline for infectious disease GWAS. Figure \ref{fig:simulationResults}A shows that our method incorporating phylogenetic information can more accurately estimate these value compared to the trait value. To ensure a fair comparison, we scaled trait values to have the same mean, zero, as host genetic effects so as not to bias the root mean squared error (RMSE) by a constant factor. As shown in the supplemental material, we can calculate the expected RMSE using the scaled trait value across scenarios in our simulation scheme because the variance in the trait due to viral genetic effects and environmental effects is fixed. Thus, we expect the RMSE using the scaled trait value to be 0.74 across all simulation scenarios. By incorporating phylogenetic information, we can improve upon this error in scenarios where the trait is highly heritable, under low selection pressure, and with relatively moderate stochastic fluctuations compared to outbreak duration. Figure \ref{fig:simulationTraitExample} gives some intuition for how this correction works by contrasting simulated scenarios with high and low heritability and low selection strength/ low stochastic fluctuations. These parameter combinations mean trait values are phylogentically correlated at clustered tips in the phylogeny. This information is the basis from which our method estimates the pathogen and corresponding non-pathogen effects on the trait. Figure \ref{fig:tip-trait-correlation} shows in more detail how trait values are correlated at closely clustered tips when an infectious disease trait is highly heritable and selection strength is low/ stochastic fluctuations are low.

@@

\begin{figure}[H]
	\begin{center}
		\includegraphics[width=\linewidth]{figures/simulation_results}
		\caption{Results from the simulation study. We simulated host, pathogen, and environmental effects on a trait under the POUMM with different heritability (y-axis) and selection strength (x-axis) parameters. For each simulated dataset, we applied our method to estimate the non-pathogen effects and performed GWAS with these values. (A) shows the root mean squared error (RMSE) of our estimator (left) compared to un-corrected trait values, scaled by their mean (right) under each simulated evolutionary scenario. The RMSE is with reference to the true (simulated) host part of the trait values. Thus, more accurate estimates (lower RMSE) mean the trait value used for GWAS will be closer to the true host part of the trait value. (B) shows how GWAS power can improve given the true, simulated non-pathogen effect on spVL (left) and using our estimate for this value (middle) compared to using the scaled trait value (right). Each tile's color corresponds to the average value across 20 simulated datasets of 500 samples. The points highlight specific heritability and selection strength values from the \emph{A. thaliana}-\emph{X. arboricola} QDR analysis, HIV-1 spVL analysis, and four simulated scenarios that are presented in more detail in Figure \ref{fig:tip-trait-correlation}. RMSE = Root mean square error, TPR = True positive rate, QDR = quantitative disease resistance, spVL = set-point viral load.}
		\label{fig:simulationResults}
	\end{center}
\end{figure}

\begin{figure}[H]
	\begin{center}
		\includegraphics[width=\linewidth]{figures/example_simulation_trait_values}
		\caption{Simulated data from two evolutionary scenarios where a phylogenetic correction to trait values improves GWAS power (right side) and where it does not (left side). These examples correspond to two of the empty points in Figure \ref{fig:simulationResults}. (A) and (B) show total trait values for 12 randomly selected tips from the simulated phylogeny with pathogen heritability 15 and 75\%, respectively. Depending on the pathogen heritability, trait values are more or less correlated at clustered tips. (C) compares our method's estimate for the non-pathogen part of trait values (y-axis) with true simulated host trait values (x-axis) with pathogen heritability 15 and 75\%. The solid line is the y=x line. Selection strength was fixed to 0.1 time $^{-1}$ for both scenarios and all other parameters were fixed as in the full simulation study.}
		\label{fig:simulationTraitExample}
	\end{center}
\end{figure}

\begin{figure}[H]
\begin{center}
\includegraphics[width = \linewidth]{figures/tip_trait_correlation.png}
	\caption{Correlations between trait values in pairs of tips in four simulated scenarios. These examples correspond to the four empty points in Figure \ref{fig:simulationResults}. Correlations are calculated for pairs of tips binned by phylogenetic distance (into deciles) across the 20 replicate simulations for each of the four evolutionary scenarios. Trait values are only noticeably correlated for closely clustered tips under the scenario with high pathogen heritability and low selection strength/ low stochastic fluctuations (upper left facet).}
	\label{fig:tip-trait-correlation}
	\end{center}
\end{figure}

@@

We applied our framework to empirical data from two different host-pathogen systems. First, we used data collected by the Swiss HIV Cohort Study (SHCS) from 1,392 individuals in Switzerland infected with HIV-1 subtype B between 1994 and 2018. The SHCS provided viral load measurements, $pol$ gene sequences, and human genotype data for these individuals. We followed the method outlined above to estimate the pathogen and non-pathogen effects on spVL for the cohort from these data. Figure \ref{fig:spvl-calc-comparison} shows the calculated (total) spVL values, which vary between approximately 1 and 6 log copies/mL in the cohort.  We estimated spVL heritability in this cohort to be 45\% (95\% highest posterior density, HPD, 24 - 67\%) and selection strength to be 58 time$^{-1}$ (95\% HPD 19 - 95) (Figure \ref{fig:poumm-parameters}, Table \ref{tab:POUMM-parameters-our-estimates}). To put these values into the context of our simulation study, they are shown as points on Figure \ref{fig:simulationResults}. The highest expected correlation in trait values between any two tips in the HIV-1 phylogeny under the POUMM was 0.45. However, Figure \ref{fig:spvl-on-tree} shows that this trait is not obviously phylogenetically structured in the cohort in general, despite high heritability. Finally, figure \ref{fig:trait-comparison} shows that the estimated non-pathogen effects on spVL correlate quite strongly with total spVL.

We compared our proposed GWAS framework with a more standard approach by performing two different GWAS on the same SHCS human genotypes. In the (i) ``GWAS with standard trait value'' we used the total trait value, our calculated spVL values, as the GWAS response variable. In the (ii) ``GWAS with estimated non-pathogen part of trait'' we used our estimates for the non-pathogen effects on spVL. Figure \ref{fig:gwas-results}A shows that results are qualitatively similar between the two GWAS. Q-Q plots show the distribution of p-values are very similar as well (Figure \ref{fig:qq-plots}). Figure \ref{fig:gwas-results}B shows how the strength of association changed for some variants in the MHC and \emph{CCR5} regions. Taking into account phylogenetic information slightly decreased association strength for most variants in the \emph{CCR5} region. Association strength increased for some variants in the MHC, for example, SNP rs9265880 had the greatest increase in significance in the MHC region, from a p-value of $3.5 \times 10^{-07}$ to $7.7 \times 10^{-09}$. However, the top-associated variants in the MHC and \emph{CCR5} regions were consistent regardless of the GWAS response variable used (Table \ref{tab:gwas-top-associations}). Finally Table \ref{tab:comp-gwas-mclaren-snps} shows how our GWAS results compare for the two top-associated SNPs identified by \citet{McLaren2015}, who performed the largest standard GWAS for HIV spVL to date. In summary, there are no clear patterns that point to new regions of association in the human genome with spVL when we take into account the pathogen phylogeny.

@@

Next, we applied our method to data collected from the \emph{A. thaliana}-\emph{X. arboricola} pathosystem by \cite{Wang2018Two-wayGenomes}. \cite{Wang2018Two-wayGenomes} performed a fully-crossed experiment in which they infected genetically diverse \emph{A. thaliana} accessions with genetically diverse strains of the phytopathogenic bacteria \emph{X. arboricola}. They scored quantitative disease resistance (QDR) on a scale of 0 (resistant) to 4 (susceptible) for up to four infected leaves for three replicates of each \emph{A. thaliana}-\emph{X. arboricola} pairing. Figure \ref{fig:qdr-on-tree} shows the inferred \emph{X. arboricola} pathogen phylogeny. Mean QDR calculated across all \emph{A. thaliana} pairings and all replicates for each \emph{X. arboricola} strain was generally low, varying between 0.11 for strain NL\_P126 and 0.78 for strain FOR\_F21. Fitting the POUMM yielded very low selection strength $\alpha$ and intensity of stochastic fluctuations  $\sigma$ parameter estimates (posterior mean 0.03 with 95\% HPD 0.0 - 0.05 and 0.03 with 95\% HPD 0.0 - 0.06, respectively; Table \ref{tab:POUMM-parameters-our-estimates-qdr}). These values deviated significantly from the respective priors (Figure \ref{fig:poumm-parameters-qdr}). Heritability, on the other hand, was quite uncertain (posterior mean 0.33 with 95\% HPD 0.0 - 0.77; Table \ref{tab:POUMM-parameters-our-estimates-qdr}). The posterior mean selection strength and heritability values are also shown in the context of the simulation study as points on Figure \ref{fig:simulationResults}.

@@

\section*{Discussion}
% State the importance of your findings
In this paper, we presented a new phylogeny-aware GWAS framework to correct for heritable pathogen effects on infectious disease traits. By using information from the pathogen phylogeny, we show it is possible to improve GWAS power to detect host genetic variants associated with a disease trait. This should help us better understand which host factors are broadly protective against a disease versus which increase susceptibility or disease severity.

@@

The main novelty of our approach is to use these POUMM parameter estimates to correct infectious disease trait values prior to performing GWAS, thereby estimating and removing pathogen effects. In simulations, we show that when trait heritability amongst infection partners is greater than approximately 25\%, GWAS power to detect host genetic variants associated with the same trait is reduced. Our method can correct for this effect in certain evolutionary scenarios by using information from the full pathogen phylogeny. Based on our simulation results, our method is anticipated to be very useful for disease traits that are highly heritable from donor to recipient and  maintain a high correlation between sampled individuals. In simulations, we showed this is the case when pathogen heritability is high, selection strength is low, and trait values are not subject to strong stochastic fluctuations. It is possible that traits under strong selection pressure would also be highly heritable and phylogenetically correlated, for instance antimicrobial resistance. However, such traits might violate the POUMM assumption that trait values vary as a random walk if they are caused by few mutations of strong affect. In summary, cohort-level, phylogenetically structured differences in the measured trait value are necessary for our approach to outperform state of the art methods.

Given our estimates for trait heritability and selection strength on HIV-1 spVL and \emph{A. thaliana} quantitative disease resistance to \emph{X. arboricola}, our simulation results reveal that we cannot expect a significant improvement in GWAS power (Figure \ref{fig:simulationResults}). Certain pairs of samples in the HIV-1 cohort were indeed expected to have phylogenetically correlated spvL values (maximum expected correlation between any two samples was 0.45). However, the overall effect on GWAS is small. For HIV-1 spVL, our phylogenetic correction slightly decreases p-values for variants in $CCR5$ and slightly decreases some and increases other p-values for variants in the MHC (Figure \ref{fig:gwas-results}B). Simulations show we shouldn't expect a net p-value decrease, but our simulations represent an ideal scenario since we simulate under the POUMM. In real life, un-modeled evolutionary pressures like drug treatment and host-specific HLA alleles might cause the reduced p-values. However, the overall picture is consistent between the two GWAS (Figure \ref{fig:gwas-results}A). Therefore, we conclude that GWAS for host determinants of HIV-1 subtype B spVL is robust to our correction for pathogen effects. For \emph{A. thaliana} QDR, inferred phylogenetic correlations are too weak to justify our phylogenetic correction (maximum expected correlation between strains was $3.2 \times 10^{-12}$). \cite{Wang2018Two-wayGenomes} found no significant \emph{A. thaliana} variants associated with QDR using a linear mixed model jointly accounting for host genetic effects, pathogen genetic effects, and interaction effects. As with HIV-1 spVL, our results do not challenge this previous finding.

Our approach is envisioned to be complementary to previously developed methods for infectious disease GWAS. First, it provides additional information on the evolutionary dynamics of the trait in the pathogen population. Then, it is a convenient pre-processing step for GWAS because it simply produces an alternate response variable for GWAS association tests. As in the \emph{A. thaliana}-\emph{X. arboricola} application, fitting the POUMM may reveal that expected phylogenetic correlations between samples are not strong enough to justify using our method to correct trait values in a GWAS. In this case, one may wish to use a linear mixed model as in \citep{Wang2018Two-wayGenomes}, where the pathogen effect is co-estimated as a random effect. In cases where a correction can be estimated and applied, it is still possible to use any of the number of previously developed GWAS models for the actual association testing (we used a linear model approach implemented in PLINK \citep{Chang2015}, though a more advanced method would be to use a linear mixed model with host ancestry as a random effect). Finally, additional model complexity can be added to the GWAS association tests. For instance, our method does not account for co-infection, which might add additional variance to trait values and decrease GWAS power. In this case, one could add co-infection status as a covariate in the GWAS association test to account for this variable.

Our method relies on the freely available R package POUMM \citep{Mitov2017a-POUMM}, which scales to trees of up to 10,000 tips \citep{Mitov2019}. All code for the simulations and HIV spVL analysis presented in this study is available on the project GitHub at https://github.com/cevo-public/POUMM-GWAS. Future applications of our method might investigate other clinically significant disease traits and outcomes that are affected by both host and pathogen genetic factors, for instance Hepatitis B Virus-related hepatocellular carcinoma \citep{An2018HostCarcinoma}, Hepatitis C treatment success \citep{Ansari2017Genome-to-genomeVirus}, and susceptibility to or severity of certain bacterial infections, e.g. \cite{Messina2016ImpactInfections, Donnenberg2015BacterialStudies}. Transcriptomic data has also previously been modeled as an evolving phenotype using an Ornstein-Uhlenbeck model \citep{Rohlfs2014ModelingVariation}. Thus, one could also estimate pathogen effects on host gene expression. In particular, it will be of interest to understand the evolutionary dynamics of these traits in the respective pathogen populations.

@@

\subsection*{\emph{A. thaliana}-\emph{X. arboricola} data}
\emph{A. thaliana} and \emph{X. arboricola} genotyping and quantitative disease resistance (QDR) measurements are described in detail in \cite{Wang2018Two-wayGenomes}. Briefly, \cite{Wang2018Two-wayGenomes} infected different \emph{A. thaliana} host accessions with different \emph{X. arboricola} pathogen strains in a fully-crossed experimental design. They infected up to 4 leaves on each of three biological replicates for each host-pathogen pairing. Then, they scored QDR for each leaf on a scale of 0 (resistant) to 4 (susceptible). We downloaded the genotype matrix with allele dosage of 33,610 SNPs for the 22 \emph{X. arboricola} pathogen strains generated by \cite{Wang2018Two-wayGenomes} from their supplemental material. We additionally downloaded a VCF file with allele dosage of 12,883,854 SNPs for the different \emph{A. thaliana} accessions from the 1001 Genomes project \citep{Alonso-Blanco20161135Thaliana}. QDR measurements were provided directly by the  \cite{Wang2018Two-wayGenomes} authors.
